Water, Methane Depletion, and High-Altitude Condensates in the Atmosphere of the Warm Super-Neptune WASP-107\lowercase{b}

### 2017-09-25 arxiv_v2/ms.tex
arxiv

\documentclass[twocolumn]{aastex61}

@@

\title{Water, Methane Depletion, and High-Altitude Condensates in the Atmosphere of the Warm Neptune WASP-107\lowercase{b}}

@@

\begin{abstract}
The Neptune-mass exoplanet WASP-107b is an exciting target for atmosphere characterization. It has an unusually large atmospheric scale height and a small, bright host star, raising the possibility of precise constraints on its current nature and formation history.  We report the first atmospheric study of WASP-107b, a Hubble Space Telescope measurement of its near-infrared transmission spectrum.  We determined the planet's composition with two techniques: atmospheric retrieval based on the transmission spectrum and interior structure modeling based on the observed mass and radius. The interior structure models set a $3\,\sigma$ upper limit on the atmospheric metallicity of $30\times$ solar. The transmission spectrum shows strong evidence for water absorption ($6.5\,\sigma$ confidence), and we infer a water abundance consistent with expectations for a solar abundance pattern. On the other hand, methane is depleted relative to expectations (at $3\,\sigma$ confidence), suggesting a low carbon-to-oxygen ratio or high internal heat flux.  The water features are smaller than predicted for a cloudless atmosphere, crossing less than one scale height. A thick condensate layer at high altitudes (0.1 - 3 mbar) is needed to match the observations; however, we find that it is challenging for physically motivated cloud and haze models to produce opaque condensates at these pressures.  Taken together, these findings serve as an illustration of the diversity and complexity of exoplanet atmospheres. The community can look forward to more such results with the high precision and wide spectral coverage afforded by future observing facilities. 
\end{abstract}

@@

\subsubsection{Final Fit}
In our final fit, we fixed the transit parameters $a/r_s$, $i$, $p$ on the precise estimates from the Kepler light curve \citep{dai17}.  We also fixed quadratic limb darkening parameters $u_1$ and $u_2$ on predictions from a PHOENIX model for a star with effective temperature 4300 K, calculated with the \texttt{limb-darkening} package from \cite{espinoza15}.  We checked that the values we fixed were consistent with our estimates when we allowed them to vary freely.  We also checked that the uncertainty in the stellar parameters does not significantly affect the PHOENIX model predictions at the level of precision of our data.  The remaining free parameters in the fit were $t_0$, $r_p/r_s$, and the systematics parameters for the visit-long and orbit-long trends.

@@

\subsection{Spectroscopic Light Curve Fits}
We binned the spectrum into 20 spectrophotometric channels from 1.12 to 1.63 $\mu$m, shown in Figure\,\ref{fig:lc}. We fit the light curves with the \texttt{divide-white} technique, which assumes that the light curve systematics have the same morphology at all wavelengths \citep{stevenson14c, kreidberg14a}. For this method, the transit model $T_\lambda(t)$ is multiplied by the systematics vector from the white light curve fit ($F_\mathrm{white}/T_\mathrm{white}$), and rescaled by a factor $C_\lambda + V_\lambda t$.  One advantage of this approach is that it removes the star spot crossing feature, enabling us to use orbit three with no additional correction. The amplitude of the feature has no detectable wavelength dependence at the level of precision of our data.  As for the white light curve, we fixed some of the transit parameters on the estimates from \cite{dai17} and fixed the limb darkening on the PHOENIX model (coefficients listed in Table\,\ref{tab:transit_depths}. The final spectroscopic light curve fits had just three free parameters per channel: $C_\lambda$, $V_\lambda$, and $r_p/r_s$.

@@

We explored several alternative choices for the spectroscopic light curve fits, but found that none of them made a significant difference in the transmission spectrum. For example, we fit the spectroscopic light curves with the same analytic model we used for the white light curve and obtained nearly identical relative transit depths. 
%(differing by just $0.3\sigma$ on average), except with a small constant offset due to the uncorrected star-spot crossing feature. This offset does not affect our final analysis because it is marginalized in the atmospheric retrieval (see \S\,\ref{sec:retrieval}). 
We also fit for a linear limb darkening parameter rather than fixing the limb darkening on the PHOENIX model values. We found that the fitted limb darkening coefficients are consistent with the model predictions, so we opted to fix the coefficients in our final analysis because it improves the precision on the transit depths by about 10\%.  We also performed an independent data reduction and fit using K. Stevenson's pipeline and found consistent results (well within $1\,\sigma$).

@@

Based on these assumptions, we fit for envelope metallicity and core mass using an MCMC with uniform priors on both parameters.  The MCMC burned in for $10^3$ steps and then collected $4\times10^6$ samples.  The envelope metal mass fractions were converted to metallicities assuming the mean molecular weight of the metals was 18 (the value for water), using the approach of \cite{fortney13}. Figure\,\ref{fig:metal_prior} shows the results.  We find a $3\,\sigma$ upper limit on the metallicity of $30\times$ solar for the young stellar age range.  Higher metallicity envelopes are not allowed because they decrease the planet's radius below the observed value.  For the older age, the upper limit is even lower ($20\times$ solar), because planets cool and contract as they age \citep{fortney08}.  These are conservative upper limits, but realistically, the planet probably formed with a core. Assuming a $5\,M_\oplus$ core, the upper limit on metallicity is 20 (10) $M_\oplus$ for the young (old) age.

@@

We explored a few retrieval scenarios with additional complexity, including the addition of cloud patchiness \citep{line16}, a quench pressure for nitrogen and carbon species \citep[e.g.][]{morley17}, and a power law haze opacity.  We also varied the assumed planet mass by $3\,sigma$. These models did not significantly improve the fit quality, and the main retrieval results were unchanged.  We also explored cases without the $30\times$ solar metallicity prior upper limit. Without this prior, metallicities up to $100\times$ were permitted, but other parameters were not significantly changed. 
%We also tested the effect of varying the planet's surface gravity within the published FIXME $\sigma$ range, and found that our results were again minimally changed.

@@

We also performed a ``free" retrieval that varied the abundances of CH$_4$, H$_2$O and NH$_3$ with no assumption of chemical equilibrium. The $3\,\sigma$ upper limit on methane abundance is $1.2\times10^{-3}$, ruling out the the predicted methane content for a composition with solar C/O and $3\,\times$ solar metallicity (the median metallicity from the chemical equilibrium retrieval).  Based on these tests, we conclude that the atmosphere of WASP-107b is depleted in methane relative to expectations for a solar abundance pattern. By contrast, the water abundance ($6\times10^{-6} - 2\times10^{-3}$) is consistent with predictions for solar composition.

@@

These results are a first look at the atmosphere of WASP-107b. The planet is already being observed at other wavelengths, including the WFC3/G102 grism and Spitzer 3.6 and 4.5 $\mu$m channels in transit and eclipse (\emph{HST} Program GO 14916, PI J. Spake, \emph{Spitzer} Program 13052, PI M. Werner; \emph{Spitzer} Program 13167, PI L. Kreidberg).  In addition, WASP-107b is included in the \emph{JWST} Guaranteed Time Observations for the NIRISS, NIRCAM, and NIRSpec instruments\footnote{\url{https://jwst-docs.stsci.edu}}.  This spate of observing programs is sure to add to the already rich and complex picture of WASP-107b's atmosphere presented here.

@@

>>>>

\documentclass[twocolumn, trackchanges]{aastex61}

@@

\title{Water, Methane Depletion, and High-Altitude Condensates in the Atmosphere of the Warm Super-Neptune WASP-107\lowercase{b}}

@@

\begin{abstract}
The super-Neptune exoplanet WASP-107b is an exciting target for atmosphere characterization. It has an unusually large atmospheric scale height and a small, bright host star, raising the possibility of precise constraints on its current nature and formation history.  We report the first atmospheric study of WASP-107b, a Hubble Space Telescope measurement of its near-infrared transmission spectrum.  We determined the planet's composition with two techniques: atmospheric retrieval based on the transmission spectrum and interior structure modeling based on the observed mass and radius. The interior structure models set a $3\,\sigma$ upper limit on the atmospheric metallicity of $30\times$ solar. The transmission spectrum shows strong evidence for water absorption ($6.5\,\sigma$ confidence), and we infer a water abundance consistent with expectations for a solar abundance pattern. On the other hand, methane is depleted relative to expectations (at $3\,\sigma$ confidence), suggesting a low carbon-to-oxygen ratio or high internal heat flux.  The water features are smaller than predicted for a cloudless atmosphere, crossing less than one scale height. A thick condensate layer at high altitudes (0.1 - 3 mbar) is needed to match the observations; however, we find that it is challenging for physically motivated cloud and haze models to produce opaque condensates at these pressures.  Taken together, these findings serve as an illustration of the diversity and complexity of exoplanet atmospheres. The community can look forward to more such results with the high precision and wide spectral coverage afforded by future observing facilities. 
\end{abstract}

@@

\subsubsection{Final Fit}
In our final fit, we fixed the transit parameters $a/r_s$, $i$, $p$ on the precise estimates from the Kepler light curve \citep{dai17}.  We also fixed quadratic limb darkening parameters $u_1$ and $u_2$ on predictions from a PHOENIX model for a star with effective temperature 4300 K, calculated with the \texttt{limb-darkening} package from \cite{espinoza15}.  We checked that the values we fixed were consistent with our estimates when we allowed them to vary freely.  We also checked that the uncertainty in the stellar parameters does not affect the PHOENIX model predictions at the level of precision of our data: we varied the stellar effective temperature by $\pm100$ K from the published value and found that the transmission spectrum was not significantly changed.  The remaining free parameters in the fit were $t_0$, $r_p/r_s$, and the systematics parameters for the visit-long and orbit-long trends.

@@

\subsection{Spectroscopic Light Curve Fits}
We binned the spectrum into 20 spectrophotometric channels from 1.12 to 1.63 $\mu$m, shown in Figure\,\ref{fig:lc}. We fit the light curves with the \texttt{divide-white} technique, which assumes that the light curve systematics have the same morphology at all wavelengths \citep{stevenson14c, kreidberg14a}. For this method, the transit model $T_\lambda(t)$ is multiplied by the systematics vector from the white light curve fit ($F_\mathrm{white}/T_\mathrm{white}$), and rescaled by a factor $C_\lambda + V_\lambda t$.  One advantage of this approach is that it removes the star spot crossing feature, enabling us to use orbit three with no additional correction. The amplitude of the feature has no detectable wavelength dependence at the level of precision of our data.  As for the white light curve, we fixed some of the transit parameters on the best fit values from \cite{dai17}. We did not put put priors on the transit parameters because varying them simply shifts the spectrum up or down by a constant value. Since the planet's 10-bar radius is a free parameter in our analysis of the spectrum, the absolute transit depth does not affect the retrieved atmospheric properties. In addition to fixing the transit parameters, we also fixed the limb darkening on the PHOENIX model predictions (coefficients listed in Table\,\ref{tab:transit_depths}.

The final spectroscopic light curve fits had just three free parameters per channel: $C_\lambda$, $V_\lambda$, and $r_p/r_s$.

@@

We explored several alternative choices for the spectroscopic light curve fits, but found that none of them made a significant difference in the transmission spectrum. In one test, we fit the spectroscopic light curves with the same analytic model we used for the white light curve. We also tested a white light curve systematics vector $S_\mathrm{white}$ from a fit that included the star-spot crossing orbit. The results from these tests are nearly identical to our nominal transmission spectrum (differing by $<0.5\sigma$ on average), except with a small constant offset due to the uncorrected star-spot crossing feature. This offset does not affect our final analysis because it is marginalized in the atmospheric retrieval with the 10-bar radius parameter (see \S\,\ref{sec:retrieval}).

We also fit for a linear limb darkening parameter rather than fixing the limb darkening on the PHOENIX model values. We found that the fitted limb darkening coefficients are consistent with the model predictions, so we opted to fix the coefficients in our final analysis because it improves the precision on the transit depths by about 10\%.  We also performed an independent data reduction and fit using K. Stevenson's pipeline and found consistent results (well within $1\,\sigma$).

@@

Based on these assumptions, we fit for envelope metallicity and core mass using an MCMC with uniform priors on both parameters.  The MCMC burned in for $10^3$ steps and then collected $4\times10^6$ samples.  The envelope metal mass fractions were converted to metallicities assuming the mean molecular weight of the metals was 18 (the value for water), using the approach of \cite{fortney13}. Figure\,\ref{fig:metal_prior} shows the results.  We find a $3\,\sigma$ upper limit on the metallicity of $30\times$ solar for the young stellar age range.  Higher metallicity envelopes are not allowed because they decrease the planet's radius below the observed value.  For the older age, the upper limit is even lower ($20\times$ solar), because planets cool and contract as they age \citep{fortney08}.  The largest sources of uncertainty in the metallicity estimate are the unknown core mass and stellar age, which are dominant over modeling uncertainties due to the equation of state and distribution of heavy elements in the envelope \citep{thorngren16}. By marginalizing over the unknown physical parameters, we obtain a conservative upper limit on the envelope metallicity.  Realistically, the planet probably formed with a core. Assuming a $5\,M_\oplus$ core, the upper limit on metallicity is 20 (10) $M_\oplus$ for the young (old) age.

@@

We explored a few retrieval scenarios with additional complexity, including the addition of cloud patchiness \citep{line16}, a quench pressure for nitrogen and carbon species \citep[e.g.][]{morley17}, and a power law haze opacity.  We also varied the assumed planet mass by $3\,\sigma$. These models did not significantly improve the fit quality, and the main retrieval results were unchanged.  We also explored cases without the $30\times$ solar metallicity prior upper limit. Without this prior, metallicities up to $100\times$ were permitted, but other parameters were not significantly changed. 
%We also tested the effect of varying the planet's surface gravity within the published FIXME $\sigma$ range, and found that our results were again minimally changed.

@@

We also performed a ``free" retrieval that varied the abundances of CH$_4$, H$_2$O and NH$_3$ with no assumption of chemical equilibrium. The $3\,\sigma$ upper limit on methane abundance is $1.2\times10^{-3}$, ruling out the the predicted equilibrium methane content at photospheric temperature and pressure (700 Kelvin, 10 mbar), for a composition with solar C/O and $3\,\times$ solar metallicity. We chose $3\times$ solar as a benchmark for comparison because this is the median metallicity from the chemical equilibrium retrieval.  Based on these tests, we conclude that the atmosphere of WASP-107b is depleted in methane relative to expectations for a solar abundance pattern. By contrast, the water abundance ($6\times10^{-6} - 2\times10^{-3}$) is consistent with predictions for solar composition.

@@

These results are a first look at the atmosphere of WASP-107b. It joins the ranks of two other relatively cool, relatively low mass exoplanets with detected water features \citep[HAT-P-11b and HAT-P-26b;][]{fraine14, wakeford17}. WASP-107b is already the target of additional observing programs at other wavelengths, including the WFC3/G102 grism and Spitzer 3.6 and 4.5 $\mu$m channels in transit and eclipse (\emph{HST} Program GO 14916, PI J. Spake, \emph{Spitzer} Program 13052, PI M. Werner; \emph{Spitzer} Program 13167, PI L. Kreidberg).  In addition, WASP-107b is included in the \emph{JWST} Guaranteed Time Observations for the NIRISS, NIRCAM, and NIRSpec instruments\footnote{\url{https://jwst-docs.stsci.edu}}.  This spate of observing programs is sure to add to the already rich and complex picture of WASP-107b's atmosphere presented here.

@@

\bibitem[{{Fraine} {et~al.}(2014){Fraine}, {Deming}, {Benneke}, {Knutson},
  {Jord{\'a}n}, {Espinoza}, {Madhusudhan}, {Wilkins}, \& {Todorov}}]{fraine14}
{Fraine}, J., {Deming}, D., {Benneke}, B., {et~al.} 2014, \nat, 513, 526

### 2017-09-25 ms.tex
arxiv

\title{Water, High-Altitude Condensates, and Methane Depletion in the Atmosphere of the Warm Neptune WASP-107\lowercase{b}}

@@

\begin{figure*}
\includegraphics[width = \textwidth]{Figures/lc.pdf}
\caption{Left: Broadband and spectrophotometric transit light curves for WASP-107b compared to best fit models. Right: residuals from the best fit. Annotations indicate the central wavelength and root-mean-square (rms) residuals. A star-spot crossing feature is shaded in gray in the upper right; our systematic error correction removes this feature from the spectroscopic light curves.} 
\label{fig:lc}
\end{figure*}

@@

\begin{figure*}
\includegraphics[width = \textwidth]{Figures/spectrum.pdf}
\caption{The transmission spectrum of WASP-107b (points with 1\,$\sigma$ error bars) compared to retrieved models (blue line with shaded $1$ and $2\sigma$ confidence intervals). The features at $1.15$ and $1.4$ $\mu$m are due to water absorption. The right-hand axis indicates the normalized atmospheric scale height assuming solar composition. The water feature crosses less than one scale height, indicating that condensates are present at high altitude.} 
\label{fig:spectrum}
\end{figure*}

@@

Based on these assumptions, we fit for envelope metallicity and core mass using an MCMC with uniform priors on both parameters.  The MCMC burned in for $10^3$ steps and then collected $4\times10^6$ samples.  The envelope metal mass fractions were converted to metallicities assuming the mean molecular weight of the metals was 18 (the value for water), using the approach of \cite{fortney13}. Figure\,\ref{fig:metal_prior} shows the results.  We find a $3\,\sigma$ upper limit on the metallicity of $30\times$ solar for the young stellar age range.  Higher metallicity envelopes are not allowed because they decrease the planet's radius below the observed value.  For the older age, the upper limit is even lower ($20\times$ solar), because planets cool and contract as they age \citep{fortney08}.  These are conservative upper limits, but realistically, the planet probably formed with a core. Assuming a $5\,M_\oplus$ core, the upper limit on metallicity is 20 (10) $M_\oplus$ for the young (old) age.

\begin{figure}
\includegraphics[width = 0.5\textwidth]{Figures/metalByAge.pdf}
\caption{Estimated envelope metallicity for a given core mass based on interior structure modeling. The red shading corresponds to the $1\sigma$ confidence interval for a young planet ($0.2-1$ Gyr); the blue is for an older age ($1-14$ Gyr).}
\label{fig:metal_prior}
\end{figure}

@@

We explored a few retrieval scenarios with additional complexity, including the addition of cloud patchiness \citep{line16}, a quench pressure for nitrogen and carbon species \citep[e.g.][]{morley17}, and a power law haze opacity.  We also varied the assumed planet mass by $3\,sigma$. These models did not significantly improve the fit quality, and the main retrieval results were unchanged.  We also explored cases without the $30\times$ solar metallicity prior upper limit. Without this prior, metallicities up to $100\times$ were permitted, but other parameters were not significantly changed. 
%We also tested the effect of varying the planet's surface gravity within the published FIXME $\sigma$ range, and found that our results were again minimally changed.

\begin{figure}
\includegraphics[width = 0.5\textwidth]{Figures/retrieval.pdf}
\caption{Retrieved distributions for (a) metallicity, (b) carbon-to-oxygen ratio, (c) irradiation temperature, and (d) cloud-top pressure in bars. Black vertical lines show the median and $\pm1\,\sigma$ confidence interval (solid and dashed, respectively). Solar metallicity and C/O values are indicated with red vertical lines.}  \label{fig:retrieval}
\end{figure}

>>>>

\title{Water, Methane Depletion, and High-Altitude Condensates in the Atmosphere of the Warm Super-Neptune WASP-107\lowercase{b}}

@@

\begin{figure*}
\includegraphics[width = \textwidth]{Figures/fig1_lc.pdf}
\caption{Left: Broadband and spectrophotometric transit light curves for WASP-107b compared to best fit models. Right: residuals from the best fit. Annotations indicate the central wavelength and root-mean-square (rms) residuals. A star-spot crossing feature is shaded in gray in the upper right; our systematic error correction removes this feature from the spectroscopic light curves.} 
\label{fig:lc}
\end{figure*}

@@

\begin{figure*}
\includegraphics[width = \textwidth]{Figures/fig2_spectrum.pdf}
\caption{The transmission spectrum of WASP-107b (points with 1\,$\sigma$ error bars) compared to retrieved models (blue line with shaded $1$ and $2\sigma$ confidence intervals). The features at $1.15$ and $1.4$ $\mu$m are due to water absorption. The right-hand axis indicates the normalized atmospheric scale height assuming solar composition. The water feature crosses less than one scale height, indicating that condensates are present at high altitude.} 
\label{fig:spectrum}
\end{figure*}

@@

Based on these assumptions, we fit for envelope metallicity and core mass using an MCMC with uniform priors on both parameters.  The MCMC burned in for $10^3$ steps and then collected $4\times10^6$ samples.  The envelope metal mass fractions were converted to metallicities assuming the mean molecular weight of the metals was 18 (the value for water), using the approach of \cite{fortney13}. Figure\,\ref{fig:metal_prior} shows the results.  We find a $3\,\sigma$ upper limit on the metallicity of $30\times$ solar for the young stellar age range.  Higher metallicity envelopes are not allowed because they decrease the planet's radius below the observed value.  For the older age, the upper limit is even lower ($20\times$ solar), because planets cool and contract as they age \citep{fortney08}.  The largest sources of uncertainty in the metallicity estimate are the unknown core mass and stellar age, which are dominant over modeling uncertainties due to the equation of state and distribution of heavy elements in the envelope \citep{thorngren16}. By marginalizing over the unknown physical parameters, we obtain a conservative upper limit on the envelope metallicity.  Realistically, the planet probably formed with a core. Assuming a $5\,M_\oplus$ core, the upper limit on metallicity is 20 (10) $M_\oplus$ for the young (old) age.

\begin{figure}
\includegraphics[width = 0.5\textwidth]{Figures/fig3_metallicity.pdf}
\caption{Estimated envelope metallicity for a given core mass based on interior structure modeling. The red shading corresponds to the $1\sigma$ confidence interval for a young planet ($0.2-1$ Gyr); the blue is for an older age ($1-14$ Gyr).}
\label{fig:metal_prior}
\end{figure}

@@

We explored a few retrieval scenarios with additional complexity, including the addition of cloud patchiness \citep{line16}, a quench pressure for nitrogen and carbon species \citep[e.g.][]{morley17}, and a power law haze opacity.  We also varied the assumed planet mass by $3\,\sigma$. These models did not significantly improve the fit quality, and the main retrieval results were unchanged.  We also explored cases without the $30\times$ solar metallicity prior upper limit. Without this prior, metallicities up to $100\times$ were permitted, but other parameters were not significantly changed. 
%We also tested the effect of varying the planet's surface gravity within the published FIXME $\sigma$ range, and found that our results were again minimally changed.

\begin{figure}
\includegraphics[width = 0.5\textwidth]{Figures/fig4_retrieval.pdf}
\caption{Retrieved distributions for (a) metallicity, (b) carbon-to-oxygen ratio, (c) irradiation temperature, and (d) cloud-top pressure in bars. Black vertical lines show the median and $\pm1\,\sigma$ confidence interval (solid and dashed, respectively). Solar metallicity and C/O values are indicated with red vertical lines.}  \label{fig:retrieval}
\end{figure}
